A Formalization of Typed Lua

### 2015-08-18 doc/papers/paper/paper.tex
small fixes

\begin{figure*}[t]
	$$
	\begin{array}{rlr}
	s ::= & & \textsc{statements:}\\
	& \;\; \mathbf{skip} & \textit{skip}\\
	& | \; s_{1} \; ; \; s_{2} & \textit{sequence}\\
	& | \; \overline{l} = el & \textit{multiple assignment}\\
	& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
	| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
	& | \; \mathbf{local} \; \overline{id{:}F} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{rec} \; id{:}F = f \; \mathbf{in} \; s & \textit{recursive function} \\
	& | \; \mathbf{return} \; el & \textit{return} \\
	& | \; \lfloor a \rfloor_{0} & \textit{application with no results}\\
	& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \textit{method declaration}\\
	e ::= & & \textsc{expressions:}\\
	& \;\; \mathbf{nil} & \textit{nil}\\
	& | \; k & \textit{other literals}\\
	& | \; id & \textit{variable access}\\
	& | \; e_{1}[e_{2}] & \textit{table access}\\
	& | \; {<}F{>} \; id & \textit{type coercion}\\
	& | \; f & \textit{function declaration}\\
	& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
	| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \textit{table constructor}\\
	& | \; e_{1} + e_{2} \;
	| \; e_{1} \; {..} \; e_{2} \;
	| \; e_{1} == e_{2} \;
	| \; e_{1} < e_{2} & \textit{binary operations}\\
	& | \; e_{1} \;\&\; e_{2} \;
	| \; e_{1} \; \mathbf{and} \; e_{2} \;
	| \; e_{1} \; \mathbf{or} \; e_{2} & \textit{binary operations}\\
	& | \; \mathbf{not} \; e \;
	| \; \# \; e & \textit{unary operations} \\
	& | \; \lfloor me \rfloor_{1} & \textit{expressions with one result}\\
	l ::= & & \textsc{left-hand values:}\\
	& \;\; id_{l} & \textit{variable assignment}\\
	& | \; e_{1}[e_{2}]_{l} & \textit{table assignment}\\
	& | \; id[k] \; {<}V{>} & \textit{type coercion}\\
	k ::= & & \textsc{literal constants:}\\
	& \;\; \mathbf{false} \; | \;
	\mathbf{true} \; | \;
	{\it int} \; | \;
	{\it float} \; | \;
	{\it string} & \\
	el ::= & & \textsc{expression lists:}\\
	& \;\; \overline{e} \; | \;
	\overline{e}, me & \\
	me ::= & & \textsc{multiple results:}\\
	& \;\; a & \textit{application}\\
	& | \; {...} & \textit{vararg expression}\\
	a ::= & & \textsc{applications:}\\
	& \;\; e(el) & \textit{function application}\\
	& | \; e{:}n(el) & \textit{method application}\\
	f ::= & & \textsc{function declarations:}\\
	& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
	pl ::= & & \textsc{parameter lists:}\\
	& \;\; \overline{id{:}F} \; | \;
	\overline{id{:}F},{...}{:}F & \\
	\end{array}
	$$
	\caption{Abstract Syntax}
	\label{fig:syntax}
\end{figure*}

@@

\begin{figure*}[t]
{\small
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-Constructor}\\
\dfrac{\begin{array}{c}
	\overline{\env_v, \penv \vdash e_1: F_1, \env_f}
	\;\;\;
\overline{\env_1, \penv \vdash e_2: F_2, \env_v}	\\
	T = \{\overline{F_1{:}vt(F_1,F_2)}\}_{unique} \;\;\;
	\env_{2} = merge(\overline{\env_{f}}) \;\;\; wf(T)
	\end{array}}
{\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-NewFieldUnique}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \{ \overline{F:V} \}_{unique}\\
	\env_{1}, \penv \vdash k:F^{new}, \env_{2} \;\;\;
	\overline{F^{new} \not\lesssim F} \;\;\;
	V^{new} = vt(F^{new}, V)
	\end{array}}
{\begin{array}{l}
	\env_{1}, \penv \vdash id[k] {<}V{>}:V^{new},
	\env_{2}[id \mapsto \{ \overline{F{:}V}, F^{new}{:}V^{new}\}_{unique}]
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-IdRead}\\
\dfrac{\env(id) = F}
{\env, \penv \vdash id:close(F), \env[id \mapsto open(F)]}
\end{array}
\begin{array}{c}
\mylabel{T-IdWrite}\\
\dfrac{\env(id) = F}
{\env, \penv \vdash id_{l}:F, \env}
\end{array}
\begin{array}{c}
\mylabel{T-IndexIdRead}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \{\overline{F:V}\} \;\;\;
	\env_{1}, \penv \vdash e_{2}:F^e, \env_{2} \;\;\;
	 F^e \lesssim F_{i}
	\end{array}}
{\env_{1}, \penv \vdash id[e_{2}]:rconst(V_{i}), \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-IndexExpRead}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash e_{1}:\{\overline{F:V}\}, \env_{2}\\
	\env_{2}, \penv \vdash e_{2}:F^e, \env_{3} \;\;\;
	 F^e \lesssim F_{i}
	\end{array}}
{\env_{1}, \penv \vdash e_{1}[e_{2}]:rconst(V_{i}), \env_{3}}
\end{array}
\begin{array}{c}
\mylabel{T-CoerceClosed}\\
\dfrac{\env_{1}(id) \subtype F \;\;\; tag(F,closed)}
{\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto reopen(F)]}
\end{array}
\begin{array}{c}
\mylabel{T-CoerceFixed}\\
\dfrac{\env_{1}(id) \subtype F \;\;\; tag(F,fixed)}
{\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto F]}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-Function}\\
\dfrac{\begin{array}{c}
	closeall(\env_{1})[\overline{id \mapsto F}], \penv[\ret \mapsto S] \vdash s, \env_{2}\\
	\env_{3} = openset(\env_{1}, frv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)) \\
	\env_{4} = closeset(\env_{3}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}F}){:}S \; s:F_1 \times \ldots F_n \times \Void \rightarrow S, \env_{3}}
\end{array}
\begin{array}{c}
\mylabel{T-Local}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash el:E, \env_{2} \\
	E \lesssim F_1 \times \ldots \times F_n \times \Value{*} \\
	\env_{2}[\overline{id \mapsto F}], \penv \vdash s, \env_{3}
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id{:}F} = el \; \mathbf{in} \; s, (\env_{3} - \{\overline{id}\})[\overline{id \mapsto \env_2(id)}]}
\end{array}
\end{array}
\]
}
\caption{Typing Rules}
\label{fig:tablerules}
\end{figure*}

@@

The main differences between Core Typed Lua and the concrete
syntax are:
\begin{itemize}
\item lack of labels and goto statements, as they are incompatible
with the {\em flow typing} of Typed Lua's type system;
\item local variable declarations have an explicit block,
instead of being just statements;
\item no {\bf elseif} clause in {\bf if} statements;
\item no {\bf repeat} and {\bf for} loops, just {\bf while};
\item the only arithmetic operator is $+$, the only relational
operators are $==$ and $<$, and the only bitwise operator
is $\&$, as all other operators have typing rules identical to
one of these;
\item type coercion for evolution of table types is explicit,
with two coercion terms, the l-value $id[k] \; {<}V{>}$ for adding a field to 
a table through assignment, and the r-value ${<}F{>} \;id$ for reinterpreting
a table type to a compatible type that is expected by the context;
\item all function declarations are anonymous, so there is a
statement for declaring recursive functions;
\item function or method applications that throw away all
results or all but the first result are explicitly annotated
as such, instead this being implicit in their syntactic context;
\item identifiers that appear in l-value position are explicitly
annotated as such with an $l$ subscript.
\end{itemize}

>>>>

\begin{figure*}[t]
	$$
	\begin{array}{rlr}
	s ::= & & \textsc{statements:}\\
	& \;\; \mathbf{skip} & \textit{skip}\\
	& | \; s_{1} \; ; \; s_{2} & \textit{sequence}\\
	& | \; \overline{l} = el & \textit{multiple assignment}\\
	& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
	| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
	& | \; \mathbf{local} \; \overline{id{:}F} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{rec} \; id{:}F = e \; \mathbf{in} \; s & \textit{recursive declaration} \\
	& | \; \mathbf{return} \; el & \textit{return} \\
	& | \; \lfloor a \rfloor_{0} & \textit{application with no results}\\
	& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \textit{method declaration}\\
	e ::= & & \textsc{expressions:}\\
	& \;\; \mathbf{nil} & \textit{nil}\\
	& | \; k & \textit{other literals}\\
	& | \; id & \textit{variable access}\\
	& | \; e_{1}[e_{2}] & \textit{table access}\\
	& | \; {<}F{>} \; id & \textit{type coercion}\\
	& | \; f & \textit{function declaration}\\
	& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
	| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \textit{table constructor}\\
	& | \; e_{1} + e_{2} \;
	| \; e_{1} \; {..} \; e_{2} \;
	| \; e_{1} == e_{2} \;
	| \; e_{1} < e_{2} & \textit{binary operations}\\
	& | \; e_{1} \;\&\; e_{2} \;
	| \; e_{1} \; \mathbf{and} \; e_{2} \;
	| \; e_{1} \; \mathbf{or} \; e_{2} & \textit{binary operations}\\
	& | \; \mathbf{not} \; e \;
	| \; \# \; e & \textit{unary operations} \\
	& | \; \lfloor me \rfloor_{1} & \textit{expressions with one result}\\
	l ::= & & \textsc{left-hand values:}\\
	& \;\; id_{l} & \textit{variable assignment}\\
	& | \; e_{1}[e_{2}]_{l} & \textit{table assignment}\\
	& | \; id[e] \; {<}V{>} & \textit{type coercion}\\
	k ::= & & \textsc{literal constants:}\\
	& \;\; \mathbf{false} \; | \;
	\mathbf{true} \; | \;
	{\it int} \; | \;
	{\it float} \; | \;
	{\it string} & \\
	el ::= & & \textsc{expression lists:}\\
	& \;\; \overline{e} \; | \;
	\overline{e}, me & \\
	me ::= & & \textsc{multiple results:}\\
	& \;\; a & \textit{application}\\
	& | \; {...} & \textit{vararg expression}\\
	a ::= & & \textsc{applications:}\\
	& \;\; e(el) & \textit{function application}\\
	& | \; e{:}n(el) & \textit{method application}\\
	f ::= & & \textsc{function declarations:}\\
	& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
	pl ::= & & \textsc{parameter lists:}\\
	& \;\; \overline{id{:}F} \; | \;
	\overline{id{:}F},{...}{:}F & \\
	\end{array}
	$$
	\caption{Abstract Syntax}
	\label{fig:syntax}
\end{figure*}

@@

\begin{figure*}[t]
{\small
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-Constructor}\\
\dfrac{\begin{array}{c}
	\overline{\env_v, \penv \vdash e_1: F_1, \env_f}
	\;\;\;
\overline{\env_1, \penv \vdash e_2: F_2, \env_v}	\\
	T = \{\overline{F_1{:}vt(F_1,F_2)}\}_{unique} \;\;\;
	\env_{2} = merge(\overline{\env_{f}}) \;\;\; wf(T)
	\end{array}}
{\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-NewFieldUnique}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \{ \overline{F:V} \}_{unique}\\
	\env_{1}, \penv \vdash e:F^{new}, \env_{2} \;\;\;
	\overline{F^{new} \not\lesssim F} \;\;\;
	V^{new} = vt(F^{new}, V)
	\end{array}}
{\begin{array}{l}
	\env_{1}, \penv \vdash id[e] {<}V{>}:V^{new},
	\env_{2}[id \mapsto \{ \overline{F{:}V}, F^{new}{:}V^{new}\}_{unique}]
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-IdRead}\\
\dfrac{\env(id) = F}
{\env, \penv \vdash id:close(F), \env[id \mapsto open(F)]}
\end{array}
\begin{array}{c}
\mylabel{T-IdWrite}\\
\dfrac{\env(id) = F}
{\env, \penv \vdash id_{l}:F, \env}
\end{array}
\begin{array}{c}
\mylabel{T-IndexIdRead}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \{\overline{F:V}\} \;\;\;
	\env_{1}, \penv \vdash e_{2}:F^e, \env_{2} \;\;\;
	 F^e \lesssim F_{i}
	\end{array}}
{\env_{1}, \penv \vdash id[e_{2}]:rconst(V_{i}), \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-IndexExpRead}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash e_{1}:\{\overline{F:V}\}, \env_{2}\\
	\env_{2}, \penv \vdash e_{2}:F^e, \env_{3} \;\;\;
	 F^e \lesssim F_{i}
	\end{array}}
{\env_{1}, \penv \vdash e_{1}[e_{2}]:rconst(V_{i}), \env_{3}}
\end{array}
\begin{array}{c}
\mylabel{T-CoerceClosed}\\
\dfrac{\env_{1}(id) \subtype F \;\;\; tag(F,closed)}
{\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto reopen(F)]}
\end{array}
\begin{array}{c}
\mylabel{T-CoerceFixed}\\
\dfrac{\env_{1}(id) \subtype F \;\;\; tag(F,fixed)}
{\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto F]}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-Function}\\
\dfrac{\begin{array}{c}
	closeall(\env_{1})[\overline{id \mapsto F}], \penv[\ret \mapsto S] \vdash s, \env_{2}\\
	\env_{3} = openset(\env_{1}, frv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)) \\
	\env_{4} = closeset(\env_{3}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}F}){:}S \; s:F_1 \times \ldots F_n \times \Void \rightarrow S, \env_{3}}
\end{array}
\begin{array}{c}
\mylabel{T-Local}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash el:E, \env_{2} \\
	E \lesssim F_1 \times \ldots \times F_n \times \Value{*} \\
	\env_{2}[\overline{id \mapsto F}], \penv \vdash s, \env_{3}
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{local} \; \overline{id{:}F} = el \; \mathbf{in} \; s, (\env_{3} - \{\overline{id}\})[\overline{id \mapsto \env_2(id)}]}
\end{array}
\end{array}
\]
}
\caption{Typing Rules}
\label{fig:tablerules}
\end{figure*}

@@

The main differences between Core Typed Lua and the concrete
syntax are:
\begin{itemize}
\item lack of labels and goto statements, as they are incompatible
with the {\em flow typing} of Typed Lua's type system;
\item local variable declarations have an explicit block,
instead of being just statements;
\item no {\bf elseif} clause in {\bf if} statements;
\item no {\bf repeat} and {\bf for} loops, just {\bf while};
\item the only arithmetic operator is $+$, the only relational
operators are $==$ and $<$, and the only bitwise operator
is $\&$, as all other operators have typing rules identical to
one of these;
\item type coercion for evolution of table types is explicit,
with two coercion terms, the l-value $id[k] \; {<}V{>}$ for adding a field to 
a table through assignment, and the r-value ${<}F{>} \;id$ for reinterpreting
a table type to a compatible type that is expected by the context;
\item all function declarations are anonymous, so there is a
statement for recursive declarations;
\item function or method applications that throw away all
results or all but the first result are explicitly annotated
as such, instead this being implicit in their syntactic context;
\item identifiers that appear in l-value position are explicitly
annotated as such with an $l$ subscript.
\end{itemize}
